Intrusion Detection with Genetic Algorithms and Fuzzy Logic

### 2013-10-16 paper.tex
Moved a few sections around, and made sure the algorithms were formated nicely.

\subsection{KDD99 Data Set}
Knowledge Discovery in Databases (KDD) is the process of extracting useful and interesting information from data in order to understand patterns in the data, with a limited amount of human intervention. Data is collected from several different sources and then is analyzed in order to produce patterns or models over the data.~\cite{bc1_ecindm} The KDD99 data set is a benchmark data set that was simulated in a military network environment in 1998, and then it was formed into the KDD99 data set in 1999~\cite{6559603}. In an experiment, the researchers divide the data set into two subsets, a \emph{training set} and a \emph{testing set}. Algorithms are trained on the training set to look for patterns. These patterns are then verified when the testing set is used.~\cite{bc1_ecindm} Four different measures are then used to determine the accuracy of an algorithm, which are described in Secion~\ref{sec:PosNeg}.

@@

\mycomment{Write more here}

\section{Fuzzy Genetic Algorithm Implementation}
The focus of the research is on detecting new or unknown types of attacks in a network. The intrusion detection system is able to identify normal network activity as well as classify different attack types using a fuzzy genetic algorithm. This kind of algorithm is able to learn new attacks, and has a high detection rate. The system used is evaluated in terms of detection speed, detection rate, and false alarm rate.~\cite{6496342, 6559603}

\subsection{RLD09 Data Set}
The KDD99 data set is 14 years old, and newer attack types are not included in it because of its age. Because of this, the authors of~\cite{6496342, 6559603} created their own data set, RLD09. To create the data set, the authors captured network data from the Computer Engineering Department at King Mongkut's University of Technology Thonburi, in Bangkok, Thailand. The data has around ten million preprocessed data packets. It has 17 different types of attacks, as well as normal network activity. The attacks can be divided into denial of service attacks and probe attacks. A packet sniffer was used to get information about TCP, UDP, and ICMP headers from protocol packets. Then this information is processed into 12 features by counting connections between a source IP and destination IP. Some example features include the number of: packets, source ports, and destination ports. The following are examples of data records, where each one has 12 feature values.

\begin{itemize}
\item 21,21,15,0,21,0,0,0,0,0,0,0, attack
\item 102,2,2,0,0,1,102,0,0,0,0,0, normal
\end{itemize}

\subsection{Fuzzy Algorithm}
In order to measure the probability of an attack, a trapezoidal shape was used in the algorithm. The trapezoidal shape has four parameters: a, b, c, and d. 
\mycomment{Put in trapezoidal shape figure.} Algorithm~\ref{alg:fuzAlg} calculates the probability of being attacked.

@@

\subsection{Algorithm Overview}
The algorithm that is used in~\cite{6496342, 6559603} first randomly finds a rule. Then the rule is improved in the training phase. After that, the rules are used to classify the data into classes in the testing phase. Algorithm~\ref{alg:fuzGenAlg} describes the fuzzy genetic algorithm that is used.

@@

The fitness function to be maximized is:
\begin{equation*}
fitness function = \frac{\alpha}{A} - \frac{\beta}{B}
\end{equation*}
In the implementation, a population size of 10 was used for each generation. An individual in the population represents a possible detection rule. The two best individuals from a present generation are preserved for the next generation. The other individuals in the new generation come from single-point crossover.

\subsection{Features and Rules}
The KDD99 data set is composed of 41 features. The authors of~\cite{6496342, 6559603} used the following eight features in their system: duration, src\_bytes, num\_failed\_logins, root\_shell, num\_access\_files, srv\_count, serror\_rate, same\_srv\_rate.

duration is the length of the connection in seconds. src\_bytes is the number of bytes sent from source to destination. num\_failed\_logins is the number of failed login attempts. root\_shell returns 1 if root shell is obtained and 0 otherwise. num\_access\_files is the number of operations on access control files. srv\_count is the number of connections to the same service as the current connection in the past two seconds. serror\_rate is the percentage of connections that have ``SYN" errors.
\mycomment{What are SYN errors?}
 same\_srv\_rate is the percentage of connections to the same service.~\cite{KDD99Features}

The value of each feature is normalized to be a number between 0.0 and 7.0, and then is encoded into blocks of binary strings. See Figure~\ref{fig:fuzEncodingForFeature} for an example of a block. A rule has 12 blocks of features, and at the end of the string is the type of attack.

One record is passed into a rule. Each feature in a record is matched to one block of the rule. The parameters of each block measure the probability of an attack using the trapezoidal fuzzy rule shape. The probabilities of each block are then combined, and the average of the probabilities is compared with a threshold to determine if the record is an attack class or normal class. Then the predicted result is compared with the actual result.

\subsection{Experimental Design and Results}
\subsubsection{Experiments using Only RLD09}
The experiments that the authors of~\cite{6496342, 6559603} performed used a total of 16,000 records of normal activity and 10,500 records of attack activity. Of the attack activity, 4,000 of them were denial of service attacks and 6,500 were probe attacks.

In the first experiment, the fuzzy genetic algorithm was used to create denial of service and probe detection rules and then the rules were verified with known attack types. 10,000 records were used for the training set and all 26,500 records were used for the testing set. The two steps in the training process were to find a denial of service rule, and find a probe rule from the training set. Both of these rules were then used in the testing process to identify attacks from the testing data set. The following algorithm was used in the testing process:

\mycomment{Figure out how to put in algorithm.}

If (dos\_rule = yes or probe\_rule = yes)

   This record is an attack;

else

   This record is normal;

@@

\begin{table}
\caption{Results from Experiment 1}
\begin{tabular}{|ccccccc|} \hline
 & Attack & Normal & Total Records & FP & FN & DR\\
DoS Training & 1499 & 8501 & 10000 & 1.46 & 47.50 & 91.64\\
Probe Training & 2496 & 7504 & 10000 & 1.83 & 15.38 & 94.79\\
Testing & 10500 & 1600 & 26500 & 1.13 & 4.10 & 97.92\\
\hline\end{tabular}
\label{tab:fuzGenExp1}
\end{table}

@@

\section{Genetic Algorithm Implementation}

\subsection{Algorithm Overview}
The authors of~\cite{DBLP:journals/corr/abs-1204-1336} used a genetic algorithm to make their intrusion detection system. The system is divided into two phases: a precalculation phase and a detection phase. In the precalculation phase, a set of chromosomes are created using training data. Then this set of chromosomes is used in the detection phase for comparison. The following algorithm is used in the precalculation phase:

\mycomment{Figure out how to put in algorithm.}

Algorithm : Initialize chromosomes for comparison

Input : Network audit data (for training)

Output : A set of chromosomes

Range = 0.125

For each training data

   If it has neighboring chromosome within Range

      Merge it with the nearest chromosome

   Else

      Create new chromosome with it

   End if

End for

In the detection phase, a population is created for test data and then the type of the data is predicted. The set of chromosomes that was created in the precalculation phase is used in the detection phase to find the fitness of each chromosome in the population. The following algorithm is used in the detection phase:

\mycomment{Figure out how to put in algorithm.}

Algorithm : Predict data/intrusion type (using GA)

Input : Network audit data (for testing), Precalculated set of chromosomes

Output : Type of data.

Initialize the population

CrossoverRate = 0.15, MutationRate = 0.35

While number of generation is not reached

   For each chromosome in the population

      For each precalculated chromosome

         Find fitness

      End for

      Assign optimal fitness as the fitness of that chromosome

   End for

   Remove some chromosomes with worse fitness

   Apply crossover to the selected pair of chromosomes of the population

   Apply mutation to each chromosome of the population

End while

>>>>

\subsection{Rules}
Rules are represented by IF-THEN statements with the following format:

IF (cond1) AND...AND (condm) THEN class = ci

\mycomment{Write more here and figure out how to format the rule}

\subsection{KDD99 Data Set}
Knowledge Discovery in Databases (KDD) is the process of extracting useful and interesting information from data in order to understand patterns in the data, with a limited amount of human intervention. Data is collected from several different sources and then is analyzed in order to produce patterns or models over the data.~\cite{bc1_ecindm}

The KDD99 data set is a benchmark data set that was simulated in a military network environment in 1999~\cite{6559603}. It is composed of 41 \emph{features}, which are things that help to distinguish normal connections from attacks.

In an experiment, the researchers divide the data set into two subsets, a \emph{training set} and a \emph{testing set}. Algorithms are trained on the training set to look for patterns. These patterns are then verified when the testing set is used.~\cite{bc1_ecindm} Four different measures are then used to determine the accuracy of an algorithm, which are described in Secion~\ref{sec:PosNeg}.

@@

\mycomment{Write more here about True Pos, True Neg, Detection Rate}

\section{Fuzzy Genetic Algorithm Implementation}
\label{sec:fuzGenAlgImp}
\mycomment{Figure out better title for Section~\ref{sec:fuzGenAlgImp}}

\subsection{Setup}
The authors of~\cite{6496342, 6559603} used 8 out of the 41 KDD99 features in their system: duration, src\_bytes, num\_failed\_logins, root\_shell, num\_access\_files, srv\_count, serror\_rate, same\_srv\_rate.

Duration is the length of the connection in seconds. src\_bytes is the number of bytes sent from source to destination. num\_failed\_logins is the number of failed login attempts. root\_shell returns 1 if root shell is obtained and 0 otherwise. num\_access\_files is the number of operations on access control files. srv\_count is the number of connections to the same service as the current connection in the past two seconds. serror\_rate is the percentage of connections that have ``SYN" errors.
\mycomment{What are SYN errors?}
 same\_srv\_rate is the percentage of connections to the same service.~\cite{KDD99Features}

\subsection{Fuzzy Algorithm}
The focus of~\cite{6496342, 6559603} is on detecting new or unknown types of attacks in a network. The intrusion detection system used is able to identify normal network activity as well as attacks using a fuzzy genetic algorithm. This kind of algorithm is able to learn new attacks, and has a high detection rate. The system used is evaluated in terms of detection speed, detection rate, and false alarm rate.

In order to measure the probability of an attack, a trapezoidal shape was used in the algorithm. The trapezoidal shape has four parameters: a, b, c, and d. 
\mycomment{Put in trapezoidal shape figure.} Algorithm~\ref{alg:fuzAlg} calculates the probability of being attacked.

@@

The four parameters a, b, c, and d are encoded into blocks of binary strings, where each block is a feature with parameters between 0.0 and 7.0. See Figure~\ref{fig:fuzEncodingForFeature} for an example of a block. A rule has 12 blocks of features, and at the end of the string is the type of attack.

One record (either an attack or normal activity) is passed into a rule. Each feature in a record is matched to one block of the rule. The parameters of each block measure the probability of an attack using the trapezoidal fuzzy rule shape. The probabilities of each block are then combined by taking the average of the probabilities and comparing them with a threshold to determine if the record is an attack class or normal class.

\subsection{Algorithm Overview}
The algorithm that is used in~\cite{6496342, 6559603} first randomly generates a rule. Then the rule is improved in the training phase. After that, the rules are used to classify the data into classes in the testing phase. Algorithm~\ref{alg:fuzGenAlg} describes the fuzzy genetic algorithm that is used.

@@

The fitness function to be maximized is:
\begin{equation*}
\textrm{fitness function} = \frac{\alpha}{A} - \frac{\beta}{B}
\end{equation*}
In the implementation, a population size of 10 was used for each generation. An individual in the population represents a possible detection rule. The two best individuals from a present generation are preserved for the next generation. The other individuals in the new generation come from single-point crossover.

\subsubsection{RLD09 Data Set}
The KDD99 data set is 14 years old, and newer attack types are not included in it because of its age. Because of this, the authors of~\cite{6496342, 6559603} created their own data set, RLD09, to use in their experiments. To create the data set, the authors captured network data from the Computer Engineering Department at King Mongkut's University of Technology Thonburi, in Bangkok, Thailand. The data has around ten million preprocessed data packets. It has 17 different types of attacks, as well as normal network activity. The attacks can be divided into denial of service attacks and probe attacks. A packet sniffer was used to get information about TCP, UDP, and ICMP headers from protocol packets. Then this information was processed into 12 features. The features are the number of TCP, UDP, and ICMP packets, the number of TCP and UDP source ports, the number of TCP and UDP destination ports, and the number of TCP fin flags, syn flags, push flags, ack flags, urgent flags.

\subsubsection{Experiments using Only RLD09}
The experiments that the authors of~\cite{6496342, 6559603} performed used a total of 16,000 records of normal activity and 10,500 records of attack activity. Of the attack activity, 4,000 of them were denial of service attacks and 6,500 were probe attacks.

In the first experiment, the fuzzy genetic algorithm was used to create denial of service and probe detection rules and then the rules were verified with known attack types. 10,000 records were used for the training set and all 26,500 records were used for the testing set. The two steps in the training process were to find a denial of service rule, and find a probe rule from the training set. Both of these rules were then used in the testing process to identify attacks from the testing data set. Algorithm~\ref{alg:testProc} was used in the testing process.

\begin{algorithm}
\caption{}
\label{alg:testProc}
\begin{algorithmic}
\IF{dos\_rule = yes or probe\_rule = yes}
  \STATE{This record is an attack;}
\ELSE \STATE{This record is normal;}
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

\begin{table*}
\caption{Results from Experiment 1}
\begin{tabular}{|ccccccc|} \hline
 & Attack & Normal & Total Records & FP & FN & DR\\
DoS Training & 1499 & 8501 & 10000 & 1.46 & 47.50 & 91.64\\
Probe Training & 2496 & 7504 & 10000 & 1.83 & 15.38 & 94.79\\
Testing & 10500 & 1600 & 26500 & 1.13 & 4.10 & 97.92\\
\hline\end{tabular}
\label{tab:fuzGenExp1}
\end{table*}

@@

\section{Genetic Algorithm Implementation}
\label{sec:genAlgImp}
\mycomment{Figure out better title for Section~\ref{sec:genAlgImp}}

\subsection{Algorithm Overview}
The authors of~\cite{DBLP:journals/corr/abs-1204-1336} used a genetic algorithm to make their intrusion detection system. The system is divided into two phases: a precalculation phase and a detection phase. In the precalculation phase, a set of chromosomes are created using training data. Then this set of chromosomes is used in the detection phase for comparison. Algorithm~\ref{alg:precalc} is used in the precalculation phase.

\begin{algorithm}
\caption{Major steps in precalculation}
\label{alg:precalc}
\begin{algorithmic}
\STATE{range = 0.125}
\FOR{each training data}
  \IF{it has neighboring chromosome within range}
    \STATE{Merge it with the nearest chromosome}
  \ELSE \STATE{Create new chromosome with it}
  \ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}

In the detection phase, a population is created for test data and then the type of the data is predicted. The set of chromosomes that was created in the precalculation phase is used in the detection phase to find the fitness of each chromosome in the population. Algorithm~\ref{alg:detection} is used in the detection phase.

\begin{algorithm}
\caption{Major steps in detection}
\label{alg:detection}
\begin{algorithmic}
\STATE{Initialize the population}
\STATE{crossoverRate = 0.15, mutationRate = 0.35}
\WHILE{number of generation is not reached}
  \FOR{each chromosome in the population}
    \FOR{each precalculated chromosome}
      \STATE{Find fitness}
    \ENDFOR
    \STATE{Assign optimal fitness as the fitness of that chromosome}
  \ENDFOR
  \STATE{Remove some chromosomes with worse fitness}
  \STATE{Apply crossover to the selected pair of chromosomes of the population}
  \STATE{Apply mutation to each chromosome of the population}
\ENDWHILE
\end{algorithmic}
\end{algorithm}
